---
title: "eUGE - **e**lectricity **U**se **G**eneration and **E**xport"
author: "Ryan Farquharson"
date: "9 August 2018"
output:
  html_document: default
---

Being a closet greenie, and a scientist, I have been fascinated by energy and water use.  Many people decry the cost of utilities with little to no understanding of how much they use, nor how much they waste.  An almost daily ritual for me is to wander up to the water meter, then down to the rain gauge, and back up to the electricity meter and record all the numbers, by hand, into an exercise book. I have around a thousand entries. Until now, I have done little with that data.  

This presented a great oportunity to hone some skills in data exploration and visualisation, plus setting up file structures, version control and doing some simple calcualtions etc.  What is described here is not necessarily the best way to do things.  Let's say it's been a good learning experience!

### Methods

#### Meter readings:

Meter readings were recorded by hand at the end of each day, whenever possible, from May 2009 to present.  The reads recorded are:

* Date (yyyy-mm-dd)
* Peak (kwh)
* Offpeak (kwh)
* PV feedin (kwh)
* PV generation (kwh)
* PV hours (hours)

Rainfall and mains water were also recorded, and climate data can be obtained from BOM.  But we have left that out for now and will focus on electricity.

The raw meter readings were transcribed into a spreadsheet and saved as the meterreads.csv file.  For this exercise, instad of trying to transcribe all of the data, I just took the full reads that were closest to the end of each month.

Some initial plots were done which revealed a number of typographical errors.  These were corrected by going back to the exercise book and checking the data.  Unfortunately I did not save or commit the incorrect file or the plots.  You'll just have to trust me on that!  It was great having the original excercise book on hand.  If this was someone else's data, QC would have been tricky.

#### Directory structures and version control
Being the first time I've worked like this, I manually set up a project folder and subfolders for doc, data, results.  I also set up a git repository and remote master.  I then opened up a new project in Rstudio.  Note that I first saved this markdown file in the doc folder but ran into trouble with the directory structure.  The code below should solve this problem.

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

#### Data processing
First I loaded the tidyverse library and used read_csv to bring the meterreads.csv file in as a tibble.

```{r data processing, echo = FALSE, message = FALSE}
library(tidyverse)

meterreads <- read_csv("data/meterreads.csv")

```

Note that with tidyverse, the date came in as datatype <date> which came in handy down the track.

Some exploratory data analysis was done using ggplot.


```{r exploratory plots, echo = FALSE}

#plot power reads i.e. peak, offpeak and PV

ggplot(meterreads, aes(x = date)) + 
  geom_line(aes(y = peak_reading, colour = "peak")) +
  geom_line(aes(y = offpeak_reading, colour = "offpeak")) +
  geom_line(aes(y = PV_reading, colour = "PV")) +
  geom_line(aes(y = feedin_reading, colour = "PV feed-in")) +
  labs(x = "Date", y = "kwh", title = "Power reads", colour = "Read type") +
  scale_color_manual(values = c("blue","red","green","black")) +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```

So far so good.  You can see the meter reads going up over time with some bumps along the way, and if you look closely enough, some different slopes at different times.

But that hides most of the story.  What we actually need is usage - i.e. the increment between each read.

The first step to get usage is to get the diffence in each meter read from the previous reading.  To do this, I used mutate() and lag().

```{r getting usage from meter reads, echo = FALSE}

# make a new tibble called 'usage' from 'meterreads' and
# add a column called peak_usage by using mutate and lag
# mutate makes a new column called peak_usage by starting with peak_reading and subtracting a value
# this value is also from the peak_reading column, but by using lag it is from the previous row.

usage <- meterreads %>% 
  mutate(peak_usage = peak_reading - 
           lag(peak_reading, default = first(peak_reading)))

# now repeat the process for the other readings

usage <- usage %>% 
  mutate(offpeak_usage = offpeak_reading -
           lag(offpeak_reading, default = first(offpeak_reading)))

usage <- usage %>% 
  mutate(feedin_usage = feedin_reading -
           lag(feedin_reading, default = first(feedin_reading)))

usage <- usage %>% 
  mutate(PV_usage = PV_reading -
           lag(PV_reading, default = first(PV_reading)))

usage <- usage %>% 
  mutate(PVhours_usage = PVhours_reading -
           lag(PVhours_reading, default = first(PVhours_reading)))

usage <- usage %>% 
  mutate(water_usage = water_reading -
           lag(water_reading, default = first(water_reading)))

usage <- usage %>% 
  mutate(days = as.integer(date -
           lag(date, default = first(date))))

usage <- slice(usage, -1)

usage <- slice(usage, -110)

usage <- slice(usage, -109)

```

There was probably a quicker way to do this, but for now, it did the trick.  One problem though, it gives us a bunch of zeros in the first row.  ggplot is smart enough to ignore these zeros, but really I should clean up this data before moving on.  See the last 3 lines where I take a slice of usage except for the first row, then the last 2 rows.


So we now have the increment of power usage, generation and export between meter reads.  

But guess what.  Months have different numbers of days, and I couldn't always do a meter read at the end of the month.  So to compare between months, we need to get usage per day.

This was done by dividing each usage by the number of days in each increment.  Happily because we're playing in the tidyverse, calculating the number of days was a cinch.  If you were paying attention, you'd have seen it in the previous chunk.

```{r getting per day data from usage, echo = FALSE}
# make a new table called perday.  divide usage by number of days to get usage per day.

perday <- usage %>% 
  mutate(peak_perday = peak_usage / days)

perday <- perday %>% 
  mutate(offpeak_perday = offpeak_usage / days)

perday <- perday %>% 
  mutate(totalusage_perday = peak_perday + offpeak_perday)

perday <- perday %>% 
  mutate(feedin_perday = feedin_usage / days)

perday <- perday %>% 
  mutate(PVgen_perday = PV_usage / days)

perday <- perday %>% 
  mutate(PVhours_perday = PVhours_usage / days)

perday <- perday %>% 
  mutate(water_perday = water_usage / days)

perday <- perday %>% 
  mutate(PVefficiency = PVgen_perday / PVhours_perday)

perday <- perday %>% 
  mutate(feedinefficiency = feedin_perday / PVhours_perday)

```

Ok. Now we have something meaningful to work with.  I did some more cleaning up and made a table called eUGE3 with months, years and seasons as factors.

```{r making eUGE tables, echo = FALSE}

eUGE1 <-  
  select(perday,"date","peak_perday","offpeak_perday","totalusage_perday",
         "PVgen_perday","feedin_perday","PVhours_perday","PVefficiency", "feedinefficiency")


# got rid of other data and trimmed to give 9 complete years

monthlist1 <- c(6,7,8,9,10,11,12)
monthlist2 <- c(1,2,3,4,5,6,7,8,9,10,11,12)
monthlist3 <- c(1,2,3,4,5)
monthlist <- c(monthlist1, monthlist2, monthlist2, monthlist2, monthlist2, monthlist2, monthlist2, monthlist2, monthlist2, monthlist3)

# Make eUGE2 which has a month list

eUGE2 <- eUGE1 %>% 
  mutate(month = as.factor(monthlist))

# Make a year list
yearlist1 <- c(2009,2009,2009,2009,2009,2009,2009)
yearlist2 <- c(2010,2010,2010,2010,2010,2010,2010,2010,2010,2010,2010,2010)
yearlist3 <- yearlist2 + 1
yearlist4 <- yearlist2 + 2
yearlist5 <- yearlist2 + 3
yearlist6 <- yearlist2 + 4
yearlist7 <- yearlist2 + 5
yearlist8 <- yearlist2 + 6
yearlist9 <- yearlist2 + 7
yearlist10 <- c(2018,2018,2018,2018,2018)

yearlist <- c(yearlist1,yearlist2,yearlist3,yearlist4,yearlist5,yearlist6,yearlist7,yearlist8,yearlist9,yearlist10)

# Add year list to eUGE2

eUGE2 <- eUGE2 %>% 
  mutate(year = as.factor(yearlist))

# Make a season list

seasonlist1 <- c('Winter', 'Winter', 'Winter', 'Spring', 'Spring', 'Spring', 'Summer')
seasonlist2 <- c('Summer', 'Summer', 'Autumn', 'Autumn', 'Autumn')
seasonlist3 <- c(seasonlist2, seasonlist1)
seasonlist <- c(seasonlist1, seasonlist3, seasonlist3, seasonlist3, seasonlist3, seasonlist3, seasonlist3, seasonlist3, seasonlist3, seasonlist2)

eUGE3 <- eUGE2 %>% 
  mutate(season = as.factor(seasonlist))

```


This is about to get exciting!  Having recored all of this data over 9 years, we're about to see what it looks like.

Let's do some exploration first.

### Data exploration

Some time ago, before it became cool, and then became a money makign exercise for the wealthy, we installed a small (1 kW) photovoltaic system connected to the grid.  Now for some PV plots.

```{r PV plots, echo = FALSE}

ggplot(eUGE3, aes(x = as.factor(month), y = PVgen_perday, colour = year)) +
  geom_point() +
  labs(x = "Month", y = "kWh per day", title = "PV generation") +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA)) +
  scale_color_viridis_d()

ggplot(eUGE3, aes(x = as.factor(month), y = feedin_perday, colour = year)) +
  geom_point() +
  labs(x = "Month", y = "kWh per day", title = "PV feed-in") +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA)) +
  scale_color_viridis_d()

```

Note the seasonality.

We can look at performance by plotting PV generation of feed-in against hours of operation.

```{r PV performance, echo = FALSE}

ggplot(eUGE3, aes(x = PVhours_perday, y = PVgen_perday, colour = month)) +
  geom_point() +
  labs(x = "Hours of operation", y = "kWh generated", title = "PV generation performance") +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

ggplot(eUGE3, aes(x = PVhours_perday, y = feedin_perday, colour = month)) +
  geom_point() +
  labs(x = "Hours of operation", y = "kWh exported", title = "PV export performance") +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```

So, how are these panels performing over time?

```{r PV efficiency plots, echo = FALSE}

# Here are some PV efficiency plots


ggplot(perday, aes(x = date)) + 
  geom_line(aes(y = PVefficiency, colour = "PV generation efficiency")) +
  geom_line(aes(y = feedinefficiency, colour = "feed-in efficiency")) +
  labs(x = "Date", y = "kwh per hour", title = "PV efficiency") +
  scale_color_manual(values = c("black","blue")) +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

```

This is the amount of electricity that our teeny weeny 1 kW PV system generates per hour of daylight.  In summer we're getting around 0.3 kW per hour.  Winter looks depressing, doesn't it?  Wondering what that dip in summer is?  Read on. 

Now let's take a look at consumption.

```{r peak usage, echo = FALSE}

ggplot(eUGE3, aes(x = as.factor(month), y = peak_perday, colour = year)) +
  geom_point() +
  labs(x = "Month", y = "kWh per day", title = "Peak usage") +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA)) +
  scale_color_viridis_d()

```

Again, note the seasonality, and the complete mismatch between PV generation and usage.  

As you can see, there's quite a spread in our use of peak electricity across years, particularly over winter.  If you look really closely, you can see that in later years we used more peak than early years.  This corresponds to the arrival of two children, going part time (and hence spending more time at home) and an air conditioner.

```{r off peak usage, echo = FALSE}
# Offpeak

ggplot(eUGE3, aes(x = as.factor(month), y = offpeak_perday, colour = year)) +
  geom_point() +
  labs(x = "Month", y = "kWh per day", title = "Off-Peak usage") +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA)) +
  scale_color_viridis_d()
```

And for off-peak electricity, you can see our reliance on an off-peak heat pump in the early years, with less reliance in later years due to the RC AC.  I think this shows up better later.

So how do we compare to other households in our area?


```{r total usage, echo = FALSE}

ggplot(eUGE3, aes(x = as.factor(month), y = totalusage_perday, colour = season)) +
  geom_point() +
  labs(x = "Month", y = "kWh per day", title = "Total usage") +
  scale_color_manual(values = c("orange","green","red", "blue")) +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA)) +
  geom_hline(yintercept = 15.7, linetype =2, colour = "red") + # average for local summer
  geom_hline(yintercept = 15.5, linetype =2, colour = "orange") + # average for local autumn
  geom_hline(yintercept = 19.0, linetype =2, colour = "blue") + # average for local winter
  geom_hline(yintercept = 16.1, linetype =2, colour = "green") + # average for local spring
  geom_hline(yintercept = 12.9, linetype =4, colour = "black") # average for our household

```

This is a plot of our total usage (peak + off-peak) over the 12 months of the year, for all of the 9 years.  Note the seasonality of usage.  

The horizontal lines represent the seasonal use for the average 4 person household in our area (http://energymadeeasy.gov.au).  The black _. line represents the average for our household for the whole year.  As you can see, we do pretty well, especially considering that we live in an old, leaky, weatherboard cottage.


Now let's put everything together.
Get ready for it... this is 'eUGE' !

## eUGE

```{r visualisation using ggplot, echo = FALSE}
# Electricity overview in kwh per day

ggplot(perday, aes(x = date)) + 
  geom_line(aes(y = peak_perday, colour = "peak")) +
  geom_line(aes(y = offpeak_perday, colour = "offpeak")) +
  geom_line(aes(y = PVgen_perday, colour = "PV generation")) +
  geom_line(aes(y = feedin_perday, colour = "PV feed-in")) +
  geom_line(aes(y = totalusage_perday, colour = "total usage")) +
  labs(x = "Date", y = "kwh per day", title = "eUGE - electricity use, generation and export", colour = "Type") +
  scale_color_manual(values = c("blue","red","green","black", "orange")) +
  geom_hline(yintercept = 15.5, linetype = 2, colour = "orange") + 
  geom_hline(yintercept = 19, linetype = 2, colour = "blue") +
  geom_hline(yintercept = 12.9, linetype = 4, colour = "black") +
  theme(panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA))

# have put in some hlines to give seasonal averages for usage of 4 person households
# average household usage from energymadeeasy.gov.au.  
# 4 person household postcode 5051
# Summer 15.7; Autumn 15.5; WInter 19.0; Spring 16.1 kWh 

# calculate average total usage per day and add in an hline.
# mean(perday$totalusage_perday, na.rm = TRUE)

```

So many stories to tell!

### So what's going on?

Well, there's a bit of background to know.  

When this kicked off it was just the two of us in a small weatherboard cottage with a newish 1 kW PV system that we paid a lot of money for (could get one 5 times bigger for the same dosh today).

We run entirely on electricity.  We have a flat panel solar hot water system with offpeak electric resistance backup.  We also have a heat bank which again runs on offpeak electricity.  For a long time, this was our sole source of heating.

Then in July 2011 bubs 1 arrived, followed shortly by a reverse cycle heat pump in the living room. We've since been tweaking our mix of heat bank (offpeak) and heat pump (peak).

In January 2014 bubs 2 arrived.

We also have visitors periodically when a supplementary hot water system kicks in, running on off peak.

So with that info, you can start to interpret the data.

One other hint. Our PV panels point north.  Around summer solstice, the sun rises and sets in the south east and south west, so there are many hours when the sun is up, the PVs are producing, but it's from diffuse light rather than direct light.  That's the dip in the efficiency plots that you saw before.

### Key findings

#### Seasonality

Both energy use and generation in our household are highly seasonal.

Useage peaks in winter.

Generation peaks in summer.

Bummer, huh?

#### 4 person households - how we compare

Depsite living in a leaky 1949 weatherboard home run entirely on electricity, our total usage is less than the average local 4 person household.  Can we do better?  Hell yeah!!!  Stay tuned.  We're currently designing our energy efficient dream home, due to be constructed in 2019.  Proper orientation, double glazed uPVC windows, R6 insulation in the ceilings, R4 in the walls, internal thermal mass, blower door tested... you get the idea.


#### Take home messages

Unfortunately our electricity use and PV generation are completely out of sync.  For now, that's ok because we get a generous feed-in tarrif. Credits in summer help pay some of the winter bills.  But we're not expecting that to last long. 

It looks like taking the current house off grid would be a money sucking challenge and counterproductive.  We would need a big system to meet heating demand in winter, and then will have a massive surplus in summer without the ability to displace dirty fossil fuels.  
Given that our main electricity use is heating, alternate fuels could be considered.  Gas is avaiable but is a dirty fossil fuel (especially when you consider fugitive emissions), isn't that cheap after all, and you loose more than half the heat up the flue.  Wood could be used, but it costs, can be environmentally dubious, and requires work which we may not be up for in our twilight years.


### Next steps
More data exploration! 

I am intimately familiar with this data, but even I found things I wasn't expecting when trying different ways of visualising the data.  There's still more I'd like to try, and potentially a few more typos I need to iron out.

It was really important to realise that whilst the big scary polt contains all of the stories, it's pretty difficult for anyone but me to tease them out.  So building up the story with some clear and simple plots is really important.

It would be interesting to transcribe some of the daily data, pull in some climate data from the BOM and test out some relationships.

We can do the same for water usage and rainfall, given that we have a tank plumbed in.

The modeller in me wants to go crazy with it.  I'm resisting that urge.
 
As we plan our new energy efficient house, we are considering heating options.  Do we need to fork out big $ for a very efficient, effective and somewhat luxurious in-slab hydronic heating system?  Do we suffice with a wood oven?  What about salvaging our heat pump and heat bank be enough?  Will we even need heating at all?

Whatever we end up with, rest assured that I will continue to record our meter readings and get back to you with some interesting comparisons, no doubt.  Hopefully it won't take me another 9 years!


